Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-05-17 SCP2008/paper/metamonads.tex
minor edits continued

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
solving these problems using linear solving and LU decomposition
as our running example. Specifically, our contributions:
%\vspace*{-2pt}
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code.
    \item Use functor type sharing constraints to encode domain-specific
      knowledge.
\end{itemize}

The rest of this paper is structured as follows: The next section
introduces code generation in MetaOCaml, the problem of name
generation, and continuation-passing style (CPS) as a general
solution.  We also introduce the monad and the issues of generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
Gaussian Elimination algorithm family in completely separate,
independent modules.  We briefly discuss related work in
section~\ref{related}. We then outline the future work and conclude.
Appendices give samples of the generated code (which is available in
full at \cite{metamonadsURL}).

@@

\section{Aspects and Functors}\label{functors}
The monad represents finer-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

While the Object-Oriented Design community has acquired an extensive
vocabulary for describing modularity ideas, the guiding principles for
modular designs has not changed since they were first articulated by
Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
\cite{EWD:EWD447}: information hiding and separation of concerns.  To
apply these principles to the study of Gaussian Elimination, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.
A study of Gaussian Elimination \cite{carette04} shows that
the following variations occur:
\begin{enumerate}
\vspace*{-6pt}
    \item \textbf{Domain}: In which (algebraic) domain do the
      matrix elements belong to.  Sometimes the domains are very
      specific (e.g., $\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p$ and
      floating point numbers), while in other cases the domains
      were left generic, e.g., multivariate polynomials over a
      field.  In the roughly 85 pieces of code surveyed
      \cite{carette04} 20 different domains were encountered.
    \item \textbf{Container}: Whether the matrix is represented as
      an array of arrays, a one-dimensional array, a hash table, a
      sparse matrix, etc., and whether indexing is done in C or
      Fortran style.  Additionally, if a particular representation
      had a special mechanism for efficient row exchanges.
    \item \textbf{Output choices}: Whether just the reduced
      matrix, or additionally the rank, the determinant, and the
      pivoting matrix are to be returned. In the larger algorithm
      family, routines like Maple's
      \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
      2^5 + 2^2 = 100$ outputs.
    \item \textbf{Fraction-free}: Whether the Gaussian Elimination
        algorithm is allowed to use unrestricted division, or only
        exact (remainder-free) division.
    \item \textbf{Pivoting}: Whether to use no, 
        column-wise, or full pivoting.
    \item \textbf{Augmented Matrices}: Whether all or only some
      columns of the matrix participate in elimination.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\vspace*{-6pt}
\begin{enumerate}
    \item \textbf{Length measure}:  For stability reasons
        (numerical or coefficient growth), if a domain
      possesses
        an appropriate length measure, this is sometimes used to choose
        an ``optimal'' pivot.
    \item \textbf{Normalization and zero-equivalence}: Whether the
      arithmetic operations of the domain give normalized results, and
      whether a specialized zero-equivalence routine is to be used.
\end{enumerate}
\noindent These are separated out from the others as they are cross-cutting
concerns: in the case of the length measure, a property of the domain
will influence the pivoting method \emph{if} pivoting is to be
performed.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated. We are grateful to anonymous reviewers for many
helpful suggestions.

>>>>

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
solving these problems using linear solving and LU decomposition
as our running example. Specifically, our contributions:
%\vspace*{-2pt}
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators which 
      will catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough breakdown of the family of LU decomposition
      algorithms.
\end{itemize}

The rest of this paper is structured as follows: The next section
introduces code generation in MetaOCaml, the problem of name
generation, and continuation-passing style (CPS) as a general
solution.  We also introduce the monad and the issues of generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
LU decomposition algorithm family in completely separate,
independent modules.  We briefly discuss related work in
section~\ref{related}. We then outline the future work and conclude.
Appendices give samples of the generated code (which is available in
full at \cite{metamonadsURL}).

@@

\section{Aspects and Functors}\label{functors}
The monad represents fine-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

While the Object-Oriented Design community has acquired an extensive
vocabulary for describing modularity ideas, the guiding principles for
modular designs has not changed since they were first articulated by
Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
\cite{EWD:EWD447}: information hiding and separation of concerns.  To
apply these principles to the study of Gaussian Elimination, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.
A study of Gaussian Elimination \cite{carette04} shows that
the following variations occur:
\begin{enumerate}
%\vspace*{-6pt}
    \item \textbf{Domain}: In which (algebraic) domain do the
      matrix elements belong to.  Sometimes the domains are very
      specific (e.g., $\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p$ and
      floating point numbers), while in other cases the domains
      were left generic, e.g., multivariate polynomials over a
      field.  In the roughly 85 pieces of code surveyed
      \cite{carette04} 20 different domains were encountered.
    \item \textbf{Container}: Whether the matrix is represented as
      an array of arrays, a one-dimensional array, a hash table, a
      sparse matrix, etc., and whether indexing is done in C or
      Fortran style.  Additionally, if a particular representation
      had a special mechanism for efficient row exchanges.
    \item \textbf{Output choices}: Whether just the reduced
      matrix, or additionally the rank, the determinant, and the
      pivoting matrix are to be returned. In the larger algorithm
      family, routines like Maple's
      \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
      2^5 + 2^2 = 100$ outputs.
    \item \textbf{Fraction-free}: Whether the Gaussian Elimination
        algorithm is allowed to use unrestricted division, or only
        exact (remainder-free) division.
    \item \textbf{Pivoting}: Whether to use no, 
        column-wise, or full pivoting.
    \item \textbf{Augmented Matrices}: Whether all or only some
      columns of the matrix participate in elimination.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\vspace*{-6pt}
\begin{enumerate}
    \item \textbf{Length measure}:  For stability reasons
        (numerical or coefficient growth), if a domain
      possesses
        an appropriate length measure, this is sometimes used to choose
        an ``optimal'' pivot.
    \item \textbf{Normalization and zero-equivalence}: Whether the
      arithmetic operations of the domain give normalized results, and
      whether a specialized zero-equivalence routine is to be used.
\end{enumerate}
\noindent These are separated out from the others as they are cross-cutting
concerns: in the case of the length measure, a property of the domain
will influence the pivoting method \emph{if} pivoting is to be
performed.

@@

\subsection*{Acknowledgments}
We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml. Many helpful discussions with Walid Taha are very
appreciated.
